---
title: "Import and representation of MERFISH mouse ileum data"
author: 
  - name: Ludwig Geistlinger    
    affiliation: Center for Computational Biomedicine, Harvard Medical School
  - name: Tyrone Lee  
    affiliation: Center for Computational Biomedicine, Harvard Medical School
  - name: Jeffrey Moffitt
    affiliation: Department of Microbiology, Harvard Medical School
  - name: Robert Gentleman
    affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{Mouse ileum}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Setup

```{r, message = FALSE}
library(MerfishData)
library(ExperimentHub)
library(ggplot2)
library(grid)
```

# Data

Spatial transcriptomics protocols based on in situ sequencing or multiplexed
RNA fluorescent hybridization can reveal detailed tissue organization.
However, distinguishing the boundaries of individual cells in such data is
challenging. Current segmentation methods typically approximate cells positions
using nuclei stains.

[Petukhov et al., 2021](https://doi.org/10.1038/s41587-021-01044-w), 
describe 
[Baysor](https://github.com/kharchenkolab/Baysor),
a segmentation method, which optimizes
2D or 3D cell boundaries considering joint likelihood of transcriptional composition
and cell morphology. Baysor can also perform segmentation based on the detected
transcripts alone.

[Petukhov et al., 2021](https://doi.org/10.1038/s41587-021-01044-w), 
compare the results of Baysor segmentation
(mRNA-only) to the results of a deep learning-based segmentation
method called 
[Cellpose](https://github.com/MouseLand/cellpose) from 
[Stringer et al., 2021](https://doi.org/10.1038/s41592-020-01018-x). 
Cellpose applies a machine learning framework for the segmentation of cell
bodies, membranes and nuclei from microscopy images.

[Petukhov et al., 2021](https://doi.org/10.1038/s41587-021-01044-w) apply
Baysor and Cellpose to MERFISH data from cryosections of mouse ileum. 
The MERFISH encoding probe library was designed to target 241 genes, including
previously defined markers for the majority of gut cell types.

Def. ileum: the final and longest segment of the small intestine.

Samples were also stained with anti-Na+/K+-ATPase primary antibodies,
oligo-labeled secondary antibodies and DAPI. MERFISH measurements across
multiple fields of view and nine *z* planes were performed to provide
a volumetric reconstruction of the distribution of the targeted
mRNAs, the cell boundaries marked by Na+/K+-ATPase IF and cell
nuclei stained with DAPI.

The data was obtained from the 
[datadryad data publication](https://doi.org/10.5061/dryad.jm63xsjb2). 

This vignette demonstrates how to obtain the MERFISH mouse ileum dataset from
[Petukhov et al., 2021](https://doi.org/10.1038/s41587-021-01044-w)
from Bioconductor's [ExperimentHub](https://bioconductor.org/packages/ExperimentHub).

```{r}
eh <- ExperimentHub()
query(eh, c("MerfishData", "ileum"))
```

## Raw data

mRNA molecule data: 820k observations for 241 genes

```{r mol-data, message = FALSE, warning = FALSE}
mol.dat <- eh[["EH7543"]]
dim(mol.dat)
head(mol.dat)
length(unique(mol.dat$gene))
```

Image data: 

1. [DAPI](https://en.wikipedia.org/wiki/DAPI) stain signal:

```{r dapi-img, message = FALSE, warning = FALSE, fig.height = 10}
dapi.img <- eh[["EH7544"]]
dapi.img
plot(dapi.img, all = TRUE)
plot(dapi.img, frame = 1)
```

2. Membrane Na+/K+ - ATPase immunofluorescence signal:

While total poly(A) and DAPI staining can provide feature-rich costains suitable
for segmentation in cell-sparse tissues such as the brain, such stains are not
as useful for segmentation in cellular-dense tissues.
To address this challenge, 
[Petukhov et al., 2021](https://doi.org/10.1038/s41587-021-01044-w)
developed protocols to combine immunofluorescence (IF) of a pan-cell-type cell
surface marker, the Na+/K+-ATPase, with MERFISH.

```{r mem-img, message = FALSE, warning = FALSE, fig.height = 10}
mem.img <- eh[["EH7545"]]
mem.img
plot(mem.img, all = TRUE)
plot(mem.img, frame = 1)
```

## Segmentation

It is also possible to obtain the data in a 
[SpatialExperiment](https://bioconductor.org/packages/SpatialExperiment), 
which integrates the segmented experimental data and cell metadata, and provides
designated accessors for the spatial coordinates and the image data.

### Baysor

Obtain dataset segmented with Baysor:

```{r baysor-spe, message = FALSE}
spe.baysor <- MouseIleumPetukhov2021(segmentation = "baysor")
spe.baysor
```

Inspect dataset:

```{r baysor-spe-show}
assay(spe.baysor, "counts")[1:5,1:5]
assay(spe.baysor, "molecules")["Acsl1",5]
colData(spe.baysor)
head(spatialCoords(spe.baysor))
imgData(spe.baysor)
```

### Cellpose

Obtain dataset segmented with Cellpose:

```{r cellpose-spe, message = FALSE}
spe.cellpose <- MouseIleumPetukhov2021(segmentation = "cellpose",
                                       use.images = FALSE)
spe.cellpose
```

Inspect dataset:

```{r cellpose-spe-show}
assay(spe.cellpose, "counts")[1:5,1:5]
colData(spe.cellpose)
head(spatialCoords(spe.cellpose))
```

### Segmentation cell counts (by cell type):

Here we inspect the difference in cell counts for the both segmentation methods,
stratified by cell type label obtained from leiden clustering and annotation by
marker gene expression:

```{r fig.width = 6, fig.height = 4}
seg <- rep(c("baysor", "cellpose"), c(ncol(spe.baysor), ncol(spe.cellpose)))
ns <- table(seg, c(spe.baysor$leiden_final, spe.cellpose$leiden_final))
df <- as.data.frame(ns, responseName = "n_cells")
colnames(df)[2] <- "leiden_final"
ggplot(df, aes(
    reorder(leiden_final, n_cells), n_cells, fill = seg)) +
    geom_bar(stat = "identity", position = "dodge") +
    xlab("") +
    ylab("Number of cells") + 
    theme_bw() +
    theme(
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

# Visualization

For visualization purposes, we focus in the following on the first *z*-plane
of the membrane staining image.

```{r}
mem.img <- imgRaster(spe.baysor, image_id = "membrane")
```

## Cell metadata

Overlay cell type annotation as in Figure 6 of the 
[publication](https://doi.org/10.1038/s41587-021-01044-w).

```{r viz-cells, results = "asis", fig.height = 8, warning = FALSE}
spe.list <- list(Baysor = spe.baysor, Cellpose = spe.cellpose)
plotTabset(spe.list, mem.img)
```

## Marker gene expression

We can also overlay the individual molecules of selected marker genes such as 
the different cluster of differentiation genes assayed in the experiment: 

```{r, fig.height = 8}
gs <- grep("^Cd", unique(mol.dat$gene), value = TRUE)
ind <- mol.dat$gene %in% gs
rel.cols <- c("gene", "x_pixel", "y_pixel")
sub.mol.dat <- mol.dat[ind, rel.cols]
colnames(sub.mol.dat)[2:3] <- sub("_pixel$", "", colnames(sub.mol.dat)[2:3])
plotXY(sub.mol.dat, "gene", mem.img)
```

## Segmentation cell borders

Here, we illustrate segmentation borders for the first *z*-plane:

```{r}
poly <- metadata(spe.baysor)$polygons
poly <- as.data.frame(poly)
poly.z1 <- subset(poly, z == 1)
```

We add holes to the cell polygons:

```{r}
poly.z1 <- addHolesToPolygons(poly.z1)
```

Plot over membrane image:

```{r, fig.height = 8}
p <- plotRasterImage(mem.img)
p <- p + geom_polygon(
            data = poly.z1,
            aes(x = x, y = y, group = cell, subgroup = subid), 
            fill = "lightblue")
p + theme_void()
```

# Interactive exploration

The MERFISH mouse ileum dataset is part of the
[gallery of publicly available MERFISH datasets](https://moffittlab.connect.hms.harvard.edu/merfish/merfish_homepage.html).

This gallery consists of dedicated 
[iSEE](https://bioconductor.org/packages/iSEE) and 
[Vitessce](http://vitessce.io/) instances, published on 
[Posit Connect](https://posit.co/products/enterprise/connect/), 
that enable the interactive exploration of different segmentations,
the expression of marker genes, and overlay of
cell metadata on a spatial grid or a microscopy image.

# SessionInfo

```{r sessionInfo}
sessionInfo()
```
