---
title: "Import and representation of MERFISH mouse hypothalamus data"
author: "Ludwig Geistlinger and Robert Gentleman"
affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{Mouse hypothalamus}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Setup

```{r, message = FALSE}
library(MerfishData)
library(ExperimentHub)
library(ggpubr)
```

# Data

[paper](https://doi.org/10.1126/science.aau5324)

Def. hypothalamic preoptic region: is a part of the anterior hypothalamus that 
controls essential social behaviors and homeostatic functions.

Cell segmentation was carried out based on total polyadenylated mRNA and
DAPI nuclei costains. Combinatorial smFISH imaging was used for the 
identification and spatial expression profiling of 161 genes in 1,027,848 cells
from 36 mice (16 female, 20 male).

The data was obtained from the 
[datadryad data publication](https://doi.org/10.5061/dryad.8t8s248)

```{r}
eh <- ExperimentHub()
query(eh, c("MerfishData", "hypothalamus"))
```

Note: complementary scRNA-seq of ~31,000 cells dissociated and captured from the
preoptic region of the hypothalamus from multiple male and female mice is available
on GEO ([GSE113576](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113576)).

## Segmented data

```{r}
spe <- MouseHypothalamusMofitt2018()
spe
```

```{r}
assay(spe)[1:5,1:5]
colData(spe)
head(spatialCoords(spe))
```

Def. Bregma: The bregma is the anatomical point on the skull at which the coronal
suture is intersected perpendicularly by the sagittal suture. Used here as a 
reference point for the twelve 1.8- by 1.8-mm imaged slices along the z-axis.

The anterior position of the preoptic region is at Bregma +0.26.

```{r}
table(spatialCoords(spe)[,"z"])
```

Cell type assignment:

```{r}
table(spe$cell_class)
```

# Visualization

Visualize as in Figure 3E of the 
[paper](https://doi.org/10.1126/science.aau5324)
for six different anterior-posterior positions from a single female mouse.

```{r}
relz <- c(0.26, 0.16, 0.06, -0.04, -0.14, -0.24)
cdat <- data.frame(colData(spe), spatialCoords(spe))
cdat <- subset(cdat, cell_class != "Ambiguous")
cdat$cell_class <- sub(" [1-4]$", "", cdat$cell_class)
cdat <- subset(cdat, z %in% relz)
cdat$z <- as.character(cdat$z)
zum <- paste(0:5 * 100, "um")
names(zum) <- as.character(relz)
cdat$z <- unname(zum[cdat$z]) 
```

```{r, fig.wide = TRUE, fig.width = 10, fig.height = 10}
pal <- get_palette("simpsons", 9)
names(pal) <- c("Endothelial", "Excitatory", "OD Immature", "Astrocyte", "Mural",
                "Microglia", "Ependymal", "Inhibitory", "OD Mature")
ggscatter(cdat, x = "x", y = "y", color = "cell_class", facet.by = "z",
          shape = 20, size = 1, palette = pal) +
          guides(color = guide_legend(override.aes = list(size = 3)))
```

# Interactive exploration

[MERFISH gallery](https://ccb-rstudio-connect.hms.harvard.edu/merfish/merfish_homepage.html)
